---
title: "Untitled"
format: docx
editor: visual
---

## Retail needs to worry about who has money to spend - what has changed about who is working and earning money?

Given the research question "Retail needs to worry about who has money to spend - what has changed about who is working and earning money?"This analysis aims to dissect the dynamics of employment and income in the context of potentially shifting consumer and purchase power bases for the retail industry, especially in light of the COVID-19 pandemic. To answer the questions thoroughly, the analysis is structured to examine both individual (people level) and sector-wide (industry level) perspectives. This approach will enable a detailed exploration of how employment status, income variations, and industry-specific trends collectively impact consumer spending and the income market landscape. 

To understand how COVID-19 has affected people's earnings, a Different in Different model is employed to reveal whether individuals experienced a change in income before and after COVID. This model uses independent variables such as demographic variables (age, sex and race) along with employment status to measure the impact on income across different groups. The "Retail Trade" industry is designated as the treatment group, which allows for a direct assessment of how this sector has been impacted compared to others. We utilize the post-treatment variable to distinguish between the periods before and after COVID-19. This allows for comparisons between pre-COVID and pandemic era periods.

```{r}
if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")
library(rio)
library(dplyr)
library(tidyverse)
library(vtable)
library(fixest)

ddi <- read_ipums_ddi("cps_00001.xml")
data <- read_ipums_micro(ddi)
industry_df <- import('indnames.csv')

df_final<- data %>% inner_join( industry_df, by = c("IND"="ind")) %>%  
  filter(YEAR >= 2018,!is.na(INCTOT))
```

Visualization

```{r}
df_final$EMPSTAT <- ifelse(df_final$EMPSTAT >= 20, "Unemployed", "Employed")
unique(df_final$RACE)
#Create a plot to compare the average income between Retail Trade and Other Industries
df_final$group <- ifelse(df_final$indname == "Retail Trade", "Retail Trade", "Other Industries")


avg_income_by_group_year <- df_final %>%
  group_by(group, YEAR) %>%
  summarise(mean_income = mean(INCTOT, na.rm = TRUE)) %>%
  ungroup()

# Avergae Income Plot
ggplot(avg_income_by_group_year, aes(x = YEAR, y = mean_income, color = group)) +
  geom_line() +
  geom_vline(xintercept = 2020, linetype="dashed", color = "red")+
  labs(title = "Difference-in-Differences: Average Income by Industry",
       x = "Year",
       y = "Average Income",
       color = "Group") +
  theme_minimal()


# Calculate employment counts by group and year
employment_counts_by_group_year <- df_final %>%
  filter(EMPSTAT == "Employed") %>%
  group_by(YEAR) %>%
  summarise(count_employed = n(), .groups = 'drop') %>%
  ungroup()

# Plot
ggplot(employment_counts_by_group_year, aes(x = YEAR, y = count_employed)) +
  geom_line() +
  geom_vline(xintercept = 2020, linetype="dashed", color = "red")+
  geom_point() + # Adding points for clarity
  labs(title = "Employment Numbers Over Time by Industry",
       x = "Year",
       y = "Count of Employed Individuals",
       color = "Group") +
  theme_minimal()
```
