---
title: "Data Translation Project - Group 6"
format: html
editor: visual
authors: Han Huynh, Erin Ballar, Thomas Taylor, & Bao Do
---

## Research Question 1

How has COVID affected the health of the retail industry, as measured by employment?

### Data Cleaning

To examine the impact of COVID-19 on retail employment, we utilized data from the IPUMS Current Population Survey. The data was filtered to include individuals aged 18-62 who were in the labor force and worked in the retail industry (NAICS codes 4670-5790). Key variables were created, including a PostCOVID indicator for observations from March 2020 onward, age groups, gender, and total employment status.

```{r setup, include=FALSE}
# Load libraries
library(ipumsr)
library(rio)
library(lubridate)
library(tidyverse)
library(fixest)
library(vtable)
library(dplyr)
library(ggplot2)
library(sandwich)
library(lmtest)
library(patchwork)

# Load dataset
ddi <- read_ipums_ddi("cps_00002.xml")
data <- read_ipums_micro(ddi)
```

```{r}
# Data Cleaning
retail_data <- data %>%
  filter(IND >= 4670 & IND <= 5790, AGE >= 18 & AGE <= 62, LABFORCE == 2) %>%
  mutate(PostCOVID = ifelse(YEAR > 2020 | (YEAR == 2020 & MONTH >= 3) | (YEAR == 2018 & MONTH == 1), 1, 0),
         AgeGroup = case_when(AGE >= 18 & AGE <= 40 ~ 0, AGE >= 41 & AGE <= 62 ~ 1),
         Gender = case_when(SEX == 1 ~ 1, SEX == 2 ~ 0, TRUE ~ NA_real_),
         TotalEmployed = ifelse(EMPSTAT %in% c(10, 12), 1, 0),
         YearMonth = as.yearmon(paste(YEAR, sprintf("%02d", MONTH), "01", sep = "-")))
```

The data was then aggregated to the year-month level, with employment rates calculated as the weighted mean of the total employed using the WTFINL survey weights. This ensures the employment measures are representative of the population.

```{r}
# Regression model
regression <- retail_data %>% 
  filter(!is.na(WTFINL),!is.na(TotalEmployed)) %>% 
  group_by(YEAR, YearMonth, AgeGroup, Gender, PostCOVID) %>%
  summarize(wtemployed = weighted.mean(TotalEmployed, WTFINL, na.rm = TRUE))
```

### Distribution of the weighted employment rate

Examining the distribution of the weighted employment rate reveals a left-skewed pattern, with a concentration of values around 0.91-0.98 and a longer left tail. This suggests that while employment rates are generally high, there are some notably lower values, likely corresponding to the pandemic period. The histogram's bins show a peak frequency around 0.94-0.96, with employment rates becoming progressively less common moving further left. This skewed shape indicates that the pandemic's impact has pulled down the typical employment rate.

```{r}
# Distribution of weighted employed 
ggplot(regression, aes(x = wtemployed)) +
  geom_histogram(binwidth = 0.01, fill = "blue", color = "white") +
  labs(title = "Distribution of Weighted Mean Employment", x = "Weighted Mean Employed") +
  theme_minimal()

```

### Overall Employment Trend Plots

To investigate the impact of the COVID-19 pandemic on the health of the retail industry, we created three graphs that provide a comprehensive view of the employment trend. The first two graphs visualize the monthly trends of employment and unemployment ratios, while the third graph examines the effect of the pandemic on the retail workforce by gender.

```{r}
# Analyzing monthly trends in employment and unemployment, grouped by year-month
yearly_data <- retail_data %>%
  group_by(YEAR, YearMonth, PostCOVID) %>%
  summarize(TotalLaborForce = sum(LABFORCE %in% c(1, 2)),
            TotalEmployed = sum(EMPSTAT %in% c(10, 12)),
            TotalUnemployed = sum(EMPSTAT %in% c(20, 21)),
            .groups = 'drop')

# Monthly Trend of Employment Ratio Graph
employment_plot <- ggplot(yearly_data, aes(x = YearMonth, y = yearly_data$TotalEmployed / yearly_data$TotalLaborForce, group = 1)) +
  geom_line(color = "darkorange", size = 1) +
  geom_point(color = "steelblue2", size = 1) +
  geom_vline(xintercept = as.numeric(as.yearmon("Mar 2020")), linetype = "dashed", color = "red") +
  annotate("text", x = as.yearmon("Mar 2020"), y = 0.9, label = "March 2020", color = "darkgreen", angle = 90, vjust = -0.5, hjust = 0) +
  labs(title = "Monthly Trend of Employment Ratio", x = "Year-Month", y = "Employment Ratio") +
  theme_classic()

# Monthly Trend of Unemployment Ratio Graph
unemployment_plot <- ggplot(yearly_data, aes(x = YearMonth, y = yearly_data$TotalUnemployed / yearly_data$TotalLaborForce, group = 1)) +
  geom_line(color = "darkorange", size = 1) +
  geom_point(color = "steelblue2", size = 1) +
  geom_vline(xintercept = as.numeric(as.yearmon("Mar 2020")), linetype = "dashed", color = "red") +
  annotate("text", x = as.yearmon("Mar 2020"), y = 0.1, label = "March 2020", color = "darkgreen", angle = 90, vjust = -0.5, hjust = 0) +
  labs(title = "Monthly Trend of Unemployment Ratio", x = "Year-Month", y = "Unemployment Ratio") +
  theme_classic()

# Combining the 2 graphs
employment_plot + unemployment_plot + plot_layout(ncol = 2)

# Exploring gender-specific dynamics in the labor market
gender_specific <- retail_data %>%
  group_by(YEAR, YearMonth, PostCOVID) %>%
  summarize(
    TotalLaborForce = sum(LABFORCE %in% c(1, 2)),
    TotalEmployed = sum(EMPSTAT %in% c(10, 12)),
    TotalUnemployed = sum(EMPSTAT %in% c(20, 21)),
    MaleEmployed = sum(SEX == 1 & EMPSTAT %in% c(10, 12)),
    FemaleEmployed = sum(SEX == 2 & EMPSTAT %in% c(10, 12)),
    MaleUnemployed = sum(SEX == 1 & EMPSTAT %in% c(20, 21)),
    FemaleUnemployed = sum(SEX == 2 & EMPSTAT %in% c(20, 21))) %>%
  ungroup()

# Make gender variables into long format
gender_long <- gender_specific %>% 
  pivot_longer(c(MaleEmployed, FemaleEmployed, MaleUnemployed, FemaleUnemployed),
               names_to = "category", values_to = "Count") 

# Labor Force Indicators by Gender Graph               
ggplot(gender_long, aes(x = YearMonth, y = Count, color = category)) + 
  geom_line() +
  scale_color_manual(values = c("#1874cd", "#106e7c","#c71936", "#fa9770")) +
  geom_vline(xintercept = as.numeric(as.yearmon("Mar 2020")), linetype = "dashed", color = "red") +
  annotate("text", x = as.yearmon("Mar 2020"), y = 900, label = "March 2020", color = "darkgreen", angle = 90, vjust = -0.5, hjust = 0) +
  labs(title = "Labor Force Indicators by Gender", y = "Count", color = "Category") + 
  theme_light() 
```

The first two graphs depict the monthly trends of employment and unemployment ratios in the retail industry, respectively. The employment ratio graph shows a relatively stable trend before the pandemic, fluctuating around 0.94 - 0.96. However, in March 2020, there is a sharp decline in the employment ratio, dropping to approximately 0.82 by April 2020. This sudden decrease aligns with the onset of the COVID-19 pandemic and the implementation of lockdown measures, which significantly impacted the retail sector. Similarly, the unemployment ratio graph shows a consistently low ratio between 0.04 - 0.06 prior to the pandemic, followed by a spike reaching over 0.18 in April 2020.

Following the initial shock, both graphs show a gradual recovery in employment. The employment ratio begins to trend upward, while the unemployment ratio starts to decline. This suggests that as businesses adapted to the new circumstances and lockdown measures were eased, the retail industry started to regain some of the lost jobs. However, it is important to note that the employment ratio does not immediately return to pre-pandemic levels, indicating that the impact of COVID-19 on retail employment persisted even as the situation improved.

The third graph breaks down the labor force indicators by gender, providing more insights on how the pandemic affected employment and unemployment for males and females in the retail sector. Before the pandemic, the counts of employed males and females remained relatively stable, with males slightly exceeding females in retail workforce. This indicates that men make up a larger portion of the retail workforce. However, the graph reveals that the impact of the pandemic on employment was higher for females compared to males. The decline in female employment during the pandemic period is steeper and more sustained than the decline in male employment, suggesting that women in the retail industry were disproportionately affected by job losses. Additionally, the recovery in employment counts appears to be slower for females, indicating that women faced greater challenges in regaining employment during the recovery phase.

### Regression Models

In this regression analysis, we aim to investigate the impact of the COVID-19 pandemic on the weighted employment (wtemployed) in the retail industry while considering the effects of age group and gender. The analysis includes three models with different specifications.

Controlling for YearMonth: We control for the YearMonth variable in all three models to account for the temporal effects on employment. By including YearMonth, we capture the overall trend in employment over time, which helps isolate the impact of the COVID-19 pandemic (PostCOVID) and other variables of interest.

Interaction Terms: In models 2 and 3, we introduce interaction terms between PostCOVID and AgeGroup, as well as PostCOVID and Gender. These interaction terms allow us to examine whether the impact of the COVID-19 pandemic on employment varies across different age groups and genders. Since there is only one retail industry in the data, including interaction terms while controlling for YearMonth enables us to capture the differential effects of the pandemic on specific subgroups within the retail industry.

Log Transformation: In model 3, we apply a log transformation to the dependent variable wtemployed. This transformation is useful when dealing with skewed data, as it helps to normalize the distribution and reduce the influence of extreme values. Using log(wtemployed) as the dependent variable enables us to interpret the coefficients as percentage changes in employment for a unit change in the independent variables.

```{r}
# Regression models
model1 <- feols(wtemployed ~ PostCOVID + YearMonth, vcov = "hetero" ,data = regression)
model2 <- feols(wtemployed ~ PostCOVID*AgeGroup + PostCOVID*Gender + YearMonth, vcov = "hetero", data = regression)
model3 <- feols(log(wtemployed) ~ PostCOVID*AgeGroup + PostCOVID*Gender + YearMonth, vcov = "hetero", data = regression)
summary(model1)
summary(model2)
summary(model3)
etable(model1, model2, model3)
```

### Regression Models Interpretation

Model 1: The baseline model estimates a 0.050 reduction in the weighted employment rate after the start of COVID, which is statistically significant at the 0.1% level. This model also controls for the overall time trend, with each additional month associated with a 0.011 increase in the weighted employment rate, statistically significant at the 1% level.

Model 2: This model extends the baseline model by including interaction terms between the PostCOVID indicator and the AgeGroup and Gender variables, allowing the effect of COVID to vary by age group and gender. The model also controls for the overall time trend. The main effect of PostCOVID remains significant (at the 0.1% level), indicating a 0.055 decrease in the weighted employment rate after the start of COVID, holding age group, gender, and time trend constant. The main effects of AgeGroup and Gender are also significant (at the 0.1% level), with each additional age group associated with a 0.023 higher employment rate and being male associated with a 0.008 higher employment rate, relative to being female. However, the interaction terms are not statistically significant, suggesting that the effect of COVID on employment does not significantly differ by age group or gender. The overall time trend remains significant, with each additional month associated with a 0.011 increase in the weighted employment rate.

Model 3: Given the skewed distribution of the weighted employment rate, this model uses a log-transformed dependent variable. The log-linear specification estimates that the post-COVID period saw an overall 6.1% decline in retail employment, statistically significant at the 0.1% level, holding age group, gender, and the time trend constant. Consistent with the other models, it also finds higher employment rates for each additional age group (by 2.4%, statistically significant at the 0.1% level) and for males (by 0.8%, statistically significant at the 0.1% level). The interaction terms remain insignificant, indicating that the pandemic's effect is relatively uniform across age groups and genders.

### Limitations

When interpreting the results of this analysis, it is important to consider potential biases and limitations. For instance, the pre-post comparison assumes that the pandemic was the only major factor differentially affecting retail employment around March 2020, but there could be other time-varying confounders not captured in the data. In addition, the models account for age and gender but do not capture other potentially relevant characteristics such as race, education, or geographic location. Consequently, the parametric assumptions of the models may not fully capture the complexity of the true relationships. Lastly, sample selection bias is a potential concern in this analysis, as the data may not be perfectly representative of the entire retail industry population. If the sample overrepresents certain retail sectors or geographic regions that were more severely affected by the pandemic, the estimated impact of COVID-19 on employment may be overstated.

### Conclusion

Across all models, the PostCOVID coefficient remains negative and statistically significant at the 0.1% level, indicating a substantial decline in retail employment after the start of the pandemic. The magnitude of the effect ranges from a 0.050 reduction in the weighted employment rate in model 1 to a 6.1% decrease in model 3. The time trend (YearMonth) is positive and significant at the 1% level in all models, suggesting a gradual increase in employment over time, holding other factors constant. Models 2 and 3 introduce age group and gender as control variables, revealing that older age groups (41 - 62) and males have significantly higher employment rates compared to their respective counterparts. However, the interaction terms between PostCOVID and AgeGroup/Gender are not statistically significant, indicating that the pandemic's impact on employment is relatively uniform across these demographic groups. The use of heteroskedasticity-robust standard errors in all models strengthens the reliability of these findings by accounting for potential non-constant variance in the errors.
